---
title: "Echinoderm Analysis"
author: "Daniel Keitley"
date: "10/05/2021"
output: html_document
---


This notebook presents analysis of the [@Foster2020] sea urchin 
(*has*) dataset as part of the Gattoni et. al 2021 
manuscript. 


```{r setup, include=FALSE}


library(reticulate)
use_condaenv("rabbit-gastrulation-python",conda = "C:/anaconda/condabin/conda",required = T) 

library(ApicalOrgan2021)
source("../utils/load.R")
library(gridExtra)
library(viridis)

set.seed(42)

```

## Load data

```{r load}
echino.sce <- loadFoster2020(foster.path)

```



## Visualise UMAP
```{r}

#echino.sce <- computeUMAP(echino.sce,min_dist=0.4,n_neighbours=75)

#AO_colours <- RColorBrewer::brewer.pal(length(unique(echino.sce$AO_annotation)),"Paired")
p <- plotUMAP(echino.sce, "AO_annotation", point_size = 0.2, 
         colours = AO_colours)

ggsave(filename="echino_umap_AO_annotation.pdf",
       path=paste0(getwd(),"/../data-out/umap"),
       plot=p)

p


p <- plotUMAP(echino.sce, colour_by="stage", point_size=0.1) + 
  scale_color_brewer(name="Stage",palette="YlGnBu") + theme_void()

ggsave(filename="echino_stage_umap.pdf",
       path=paste0(getwd(),"/plots"),
       plot=p, dpi=300)


```

## Visualise 3D UMAP
```{r}
echino.3d <- computeUMAP(echino.sce,min_dist=0.4,n_neighbours=50,n_dims = 3)

p <- plot3dUMAP(echino.3d,"AO_annotation",
                colours=AO_colours)

htmlwidgets::saveWidget(p, "data-out/umap/umap3d_AO_annotation.html")



```




In this dataset we are interested in identifying the cell types involved in the 
specification of the apical organ. To identify these cells, we reannotate the 
dataset using known apical organ marker genes reported in the literature.


```{r}
plotDimredHighlight(echino.sce, dimred="UMAP",
                    colour_by="AO_annotation",
                    highlight=c("Serotonergic neurons",
                                
                                "Early AP", "Late AP"),
                    colours=echino.AO_colours,
                    point_size=0.8) + 
  theme_void() + theme(aspect.ratio=1)

```


## AO Expression
```{r}
AO_genes <- read.csv("G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/data-in/echino_heatmap_genes.tsv",
                     sep=";")

p <- plotDotPlotHeatmap(echino.sce,echino.sce$AO_heatmap_annotation,
                   echino.sce$stage,
                   AO_genes$Code,AO_genes$Gene,
                   plot_stage_legend = TRUE)

ggsave(filename="echino_AO_dotplot.pdf",
       path=paste0(getwd(),"/plots"),
       plot=p, dpi = 300, width=9.15, height=4.2, units="in")

```



```{r}

k <- 1
ps <- apply(rbind(AO_genes,c("TPH","LOC578903")),1,function(x) {
  p <- plotGeneUMAP(echino.sce, x[["Code"]], x[["Gene"]], point_size=0.2)
  k <- k + 1
  return(p)
})

p <- do.call("grid.arrange", c(ps, nrow=2))

ggsave(filename="echino_AO_genes_umap.pdf",
       path=paste0(getwd(),"/plots"),
       plot=p, dpi = 300, width=30, height=8, units="in")

```

The expression of these genes show a distinct pattern at early and late 
developmental stages.


```{r}

AP_genes <- read.csv("G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/data-in/echino_stage_heatmap_genes.tsv",
                     sep=";")

stage_exp <- getExpressionFast(echino.sce, AP_genes["Code"],group_by="stage", norm="min-max")

Heatmap(stage_exp,
        name = "Mean\nexp",
        cluster_rows = T,
        cluster_columns = F,
        row_labels = AP_genes$Gene,
        column_labels = c("8c","64c","M","EB","HB","MB","EG","LG"),
        column_names_rot = 45,
        column_split = c(rep("Early",4),rep("Late",4)),
        column_gap = unit(0, "mm"), 
        border = TRUE,
        col=circlize::colorRamp2(c(0,0.5,1),
                                 viridis(3,option="inferno")))
```




```{r}
df <- melt(table(echino.sce$stage[echino.sce$AO_annotation == "Early AP"]))
late.df <- melt(table(echino.sce$stage[echino.sce$AO_annotation == "Late AP"]))

df$celltype <- "Early AP"
late.df$celltype <- "Late AP"
df <- rbind(df,late.df)

p <- ggplot(df) + geom_col(aes(x=Var1,y=value,fill=celltype)) + 
  scale_fill_manual(values=echino.AO_colours) +
  facet_grid(~celltype) + 
  theme_void() + theme(legend.position="none",
                       strip.text.x = element_blank(),
                       aspect.ratio=1) 

ggsave(filename="echino_dotplot_hists.pdf",
       path=paste0(getwd(),"/plots"),
       plot=p, dpi = 300, width=5, height=5, units="in")


```



## Annotate AO cell types
```{r}

```


## Visualise new annotation

```{r tsne, fig.width = 9.5,fig.height=3}
p1 <- plotTSNE(echino.sce,colour_by="original_annotation",legend_position = "left")

p2 <- plotTSNEHighlight(echino.sce,colour_by="AO_annotation",
                        highlight=c("Serotonergic neurons",
                                    "Early AP", "Late AP",
                                    "Other neural"),
                        colours=echino.AO_colours)

grid.arrange(p1,p2,ncol=2)
```






## Differential expression


```{r}
celltype_comps <- list(c("Early AP", "All"),
              c("Late AP", "All"),
              c("Serotonergic neurons", "All"),
              c("Early AP", "Late AP"),
              c("Late AP", "Serotonergic neurons"),
              c("Early AP", "Other neural"),
              c("Late AP", "Other neural"),
              c("Serotonergic neurons", "Other neural"))

echino.edgeR <- prepareEdgeR(echino.sce, "AO_annotation", 
                             celltype_comps[[1]][1], celltype_comps[[1]][2],
                             block_by = "stage")

out <- runEdgeR(echino.edgeR)




#echino.sce$group[echino.sce$group %in% c("","18")] <- "9+18"
#cluster_comps <- list(c())

```





# FindMarkers
```{r}

# SU – Clust 17 vs Ectodermal (HB onwards). vs Other neural
echino.ecto <- echino.sce[,echino.sce$original_annotation %in% c("Neural","Aboral ectoderm","Oral ectoderm")
                          & echino.sce$stage %in% c("Hatched blastula","Mesenchyme blastula",
                                                    "Early gastrula","Late gastrula")]
markers <- findMarkers(echino.ecto,groups=echino.ecto$seurat_clusters,pval.type="all",
                          block=echino.ecto$orig.ident)
write.table(markers$`17`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/clust17_vs_ectoderm/degs_wblock.tsv",
            sep="\t",quote = F)


# SU – Clust 17 vs (HB onwards). Other neural
echino.neural <- echino.sce[,echino.sce$original_annotation %in% c("Neural")
                          & echino.sce$stage %in% c("Hatched blastula","Mesenchyme blastula",
                                                    "Early gastrula","Late gastrula")]
markers <- findMarkers(echino.neural,groups=echino.neural$seurat_clusters,pval.type="all",
                          block=echino.neural$orig.ident)
write.table(markers$`17`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/clust17_vs_neural/degs_wblock.tsv",
            sep="\t",quote = F)


# SU – Clust 17 vs Other Nervous
echino.neural <- echino.sce[,echino.sce$original_annotation %in% c("Neural")]
markers <- findMarkers(echino.neural,groups=echino.neural$seurat_clusters,pval.type="all",
                          block=echino.neural$orig.ident)
write.table(markers$`17`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/clust17_vs_neural/degs_wblock_all_stages.tsv",
            sep="\t",quote = F)


# SU – Clust 9 + 18 vs Other Nervous
echino.neural <- echino.sce[,echino.sce$original_annotation %in% c("Neural")]
echino.neural$DE_annotation <- as.character(echino.neural$seurat_clusters)
echino.neural$DE_annotation[echino.neural$DE_annotation %in% c("9","18")] <- "9+18"
markers <- findMarkers(echino.neural,groups=echino.neural$DE_annotation,pval.type="all",
                          block=echino.neural$orig.ident)
write.table(markers$`9+18`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/clust9+18_vs_neural/degs_wblock_all_stages.tsv",
            sep="\t",quote = F)


# SU – Clust 9 + 18 vs Ectodermal
echino.ecto <- echino.sce[,echino.sce$original_annotation %in% c("Neural","Aboral ectoderm","Oral ectoderm")]
echino.ecto$DE_annotation <- as.character(echino.ecto$seurat_clusters)
echino.ecto$DE_annotation[echino.ecto$DE_annotation %in% c("9","18")] <- "9+18"
markers <- findMarkers(echino.ecto,groups=echino.ecto$DE_annotation,pval.type="all",
                          block=echino.ecto$orig.ident)
write.table(markers$`9+18`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/clust9+18_vs_ectoderm/degs_wblock_all_stages.tsv",
            sep="\t",quote = F)


# SU - Early AP + Late AP + Serotonergic neurons vs ectoderm
echino.ecto <- echino.sce[,(echino.sce$original_annotation %in% c("Neural","Aboral ectoderm","Oral ectoderm"))
                          & (echino.sce$stage %in% c("Hatched blastula","Mesenchyme blastula",
                                                    "Early gastrula","Late gastrula"))]
echino.ecto$DE_annotation <- echino.ecto$AO_annotation
echino.ecto$DE_annotation[echino.ecto$DE_annotation %in% c("Early AP","Late AP","Serotonergic neurons")] <- "AP"
markers <- findMarkers(echino.ecto,groups=echino.ecto$DE_annotation,pval.type="all",
                          block=echino.ecto$orig.ident)
write.table(markers$`AP`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/AP_vs_ectoderm/degs_wblock_HB_onwards.tsv",
            sep="\t",quote = F)


# SU - Early AP + Late AP + Serotonergic neurons vs neural
echino.neural <- echino.sce[,(echino.sce$original_annotation %in% c("Neural"))
                          & (echino.sce$stage %in% c("Hatched blastula","Mesenchyme blastula",
                                                    "Early gastrula","Late gastrula"))]
echino.neural$DE_annotation <- echino.neural$AO_annotation
echino.neural$DE_annotation[echino.neural$DE_annotation %in% c("Early AP","Late AP","Serotonergic neurons")] <- "AP"
markers <- findMarkers(echino.neural,groups=echino.neural$DE_annotation,pval.type="all",
                          block=echino.neural$orig.ident)
write.table(markers$`AP`,"G:/My Drive/Postgrad/PhD/Projects/ApicalOrgan2021/gene_lists/differential_expression/sea urchin/AP_vs_neural/degs_wblock_HB_onwards.tsv",
            sep="\t",quote = F)









# Find differentially expressed genes
markers <- findMarkers(echino.sce,groups=echino.sce$seurat_clusters,pval.type="some",
                          block=echino.sce$orig.ident)
AO_markers <- markers$`17`


write.table(AO_markers[,1:3],"G:/My Drive/Postgrad/PhD/Projects/apical_organ/gene_lists/differential_expression/sea urchin/cluster17_degs.tsv",
          sep="\t",quote = F)


gene_names <- read.csv(paste0(foster.path,"code-gene translation table.csv"),sep=";")
gene_names = gene_names[!duplicated(gene_names$Locus),] # Remove duplicated codes
rownames(gene_names) <- gene_names$Locus

gene_names[gene_names$Locus=="foxq2","Protein.Name"] <- "foxq2"


AO_markers <- cbind(AO_markers,gene_names[rownames(AO_markers),"Protein.Name"])
colnames(AO_markers)[ncol(AO_markers)] <- "gene_desc"
AO_markers$gene_id <- rownames(AO_markers)

AO_markers <- AO_markers[,c("gene_id","gene_desc","p.value","FDR","summary.logFC")]

write.table(AO_markers[,c("gene_id","gene_desc")],"G:/My Drive/Postgrad/PhD/Projects/apical_organ/gene_lists/differential_expression/sea urchin/AO_celltypes_degs.tsv",
          sep="\t",quote = F)

highlight_genes <- c("foxq2","LOC575604","NK2.1","LOC583193","LOC591840","LOC582578","LOC575572","LOC583414","LOC579532")
highlight_gene_names <- c("foxq2","sfrp5","NK2.1","foxe3","basp1","LOC582578","fzd5","Rx1","dkk3","Hmx")

# Add edited gene name

EnhancedVolcano(AO_markers,x="summary.logFC",y="FDR",
                ylab="-log2 FDR",
                xlab="log2 FC ",
                lab=rownames(AO_markers),
                labSize = 4,
                drawConnectors = T,arrowheads = F,
                title="",subtitle = "",caption = "",
                legendPosition = "none",
                shape=16) 

```


